Epistemic Semantics in Guarded String Models

### 2020-06-15 tex/draft.tex
Section 5 in draft almost finished.

\title{Contstructive Epistemic Semantics in Guarded String Models}

@@

\section{Syntax-semantics interface}
An architecture of interpretation by transtlation is employed, where English
sentences are mapped to terms in a logical language via an interpreted grammar, and
these terms are intreted in a guarded string model as propositions. For the latter,
there are options of translation into the finite state calculus, and representation
in Haskell via lazy lists of guarded strings.

The grammar is a semantically interpreted multi-modal categorial grammar, consisting
of a lexicon of words, their categorial types, and interpretations in a logical
lambda language.  \ex{phenomena1} lists and exemplifies phenomena that are covered.

@@

\ex{lexicon1} gives illustrative lexical entries. The grammar and semantics are in
certain way optimized for the simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute the epistemic
alternative relations for those agents, rather than individuals.  This is possible
because the agents are never arguments of extensional predicates, so what matters
about the agents is their epistemic alternative relations.  The root verb {\em know}
contributes existential modal force.  The complementizers {\em that} and {\em
  whether} are the heads of their dominating clauses, and assemble an alternative
relation, modal force, and proposition contributed by the complement, in part by
introducing the dual of the existential modal force to obtain universal modal force.
These moves are bizarre, but are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated with more complicated
constructions that are better motivated in more comprehensive interpreted grammar of
English.

@@

\hspace{-10mm}\begin{tabular}[t]{l|c|c}
Amy & $e$ & $\doublehat{R}_a$\\
Bob & $e$ & $\doublehat{R}_b$\\ 
heads & $d \backslash_D t$ & $W{\cdot}h$ \\
it's & $t\slash(d \backslash_D t)$ & $\lambda p.p$ \\
\end{tabular}
\end{example}

>>>>

% This used to say Constructive ..., which is a better title.
% Reserve that for a journal paper if we get to that.
\title{Epistemic Semantics in Guarded String Models}

@@

\pagebreak
\section{Syntax-semantics interface}
An architecture of interpretation by transtlation is employed, where English
sentences are mapped to terms in the logical language (--) via an interpreted
grammar, and these terms are in turn interpreted as propositions (sets of possible
worlds). For the latter, there are options of translation into the finite state
calculus in order to represent propositions as finite state machines (Section 3), and
representation in Haskell via lazy lists of guarded strings (Section 4).  The grammar
is a semantically interpreted multi-modal categorial grammar, consisting of a lexicon
of words, their categorial types, and interpretations in a logical lambda language.
\ex{phenomena1} lists that are covered.

@@

\ex{lexicon1} gives illustrative lexical entries. The grammar and semantics are in
certain way optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute the epistemic
alternative relations for those agents, rather than individuals.  This is possible
because the agents are never arguments of extensional predicates, so what matters
about the agents is their epistemic alternative relations.  The root verb {\em know}
contributes existential modal force.  The complementizers {\em that} and {\em
  whether} are the heads of their dominating clauses, and assemble an alternative
relation, modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to arrive universal
modal force.  These moves are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated in a more comprehensive
interpreted grammar of English.

@@

\hspace{-10mm}\begin{tabular}[t]{l|c|c}
Amy & $e$ & ${R}_a$\\
Bob & $e$ & ${R}_b$\\
it & $d$ & $d$ \\
heads & $d \backslash_D t$ & $\lambda x.0^c{\cdot}h$ \\
tails & $d \backslash_D t$ & $\lambda x.0^c{\cdot}!h$ \\
is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P.\lambda x.P x$ \\
knows & $(e \backslash t)\slash_M t$ &  $\lambda p.\lambda R. {\Diamond}R p$\\
that  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
whether  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
& \multicolumn{2}{r}{$+ {\sim}(m p R)$} \\
\end{tabular}
\end{example}

Multimodal categories such as $\backslash_D$ and $\backslash_M$ are used to control
the derivation.  For instance the category of {\em heads} $d \backslash_D t$.  The
dummy expletive subject {\em it} has category $d$, but the phrase {\em it heads} of
category $t$ can not be formed, because $\backslash_D$ is not syntactically active as
a function.  Instead {\em it is heads} can be formed with a predicator {\em is} of
category $(d \backslash t)\slash(d \backslash_D t)$.  (This uses Lambek/Bar-Hillel
notation for slashes, so that $(d \backslash t)\slash(d \backslash_D t)$ combines
with $d \backslash_D t$ on the right to give a value that combines with $d$ on the
left to give $t$.)  Similarly {\em knows} has a category with the top-level slash
$\slash_M$, and combines to form a sentence as an argument of {\em that} or {\em
  whether}, which has a category that looks for the category of {\em know} on the
left, after combining with a complement sentence on the right.

The semantic translations in the third column of \ex{lexicon1} use the Epik term
language, incremented with lambda.  The body of $\lambda x.0^c{\cdot}h$, which is the
semantic lexical entry for {\em heads}, is a term denoting the set of all worlds
where the coin is heads, represented as the set of all guarded strings that end with
a Boolean valuation where the primitive proposition $h$ (it's heads) is true.  There
is $\lambda x$ at the front because of a correspondence the grammar formalism uses a
correspondence between syntactic and semantic types. However, it does not bind
anything, because sentences such as {\em it isn't heads} have an expletive subject.
The body ${\Diamond}R p$ of $\lambda p.\lambda R. {\Diamond}R p$, which is the
semantic lexical entry of {\em knows}, is an Epik term denoting the pre-image of the
set of worlds $p$ according to the relation $\doublehat{R}$ between guarded strings
that is determined by the event-level relation $R$.  This is not the right semantics
for {\em Amy knows that it's heads}, because it is an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the complementizer
{\em that} or {\em whether}, which introduces the dual.

Sentences are parsed with a chart parser for categorial grammar.  The semantics for
complex phrases are obtained by syntactic application of semantic translations,
accompanied by beta reduction.  Semantic terms in the parsing formalism are
expressions of untyped lambda calculus.  The grammar is set up so that lambda is
eliminated by beta reduction in the semantic term corresponding to a sentence. In
consequence, the semantic term translating a sentence is a term of the Epik language
(--).  Such a term designates a set of possible words (guarded strings) in the
possible worlds model deterimened by an Epik specification such as the one in Figure
1.  \exi{three}{a} is an English sentence with predicate conjunction and three levels
of clausal embedding.  Using the grammar and parser, the sentence is mapped to the
Epik term \exi{b}.  Using the result from Section 3, this term can be mapped in an
implementation of the finite state calculus to a finite state machine that represents
a countably infinite set of possible worlds, represented as guarded strings.  Using
the result from Section 4, it can be mapped to an infinite lazy list of guarded
strings, representing the same set of possible worlds.  Either of these is a concrete
computational representation of the propositional semantic value \semval{Amy knows
  that Bob knows that Amy knows whether it is heads and knows that Bob does not know
  that Amy knows that it is tails}, in the familiar sense of Montague semantics for
natural language.

\begin{examples}  \label{three}
\itema  Amy knows that Bob knows that Amy knows whether it is heads and
knows that Bob does not know that Amy knows that it is tails.
\itemb
\end{examples}
